ODGI: understanding pangenome graphs

### 2021-10-13 manuscript/main.tex
working into the body

\begin{enumerate}
\item They are relatively sparse, with low average node degree.
\item They can be sorted, so that most edges go between nodes that are close together in the sort order (and identifier space).
\item Paths are similar to each other.
%We incorporate concepts first introduced in the dynamic version of the Graph BWT \citep{Siren:2020} (GBWT), a scalable implementation of the graph extension of the positional Burrows-Wheeler transform \citep{Durbin_2014}.
\end{enumerate}

%These properties and this  encourage us to use a node-centric model for the graph, where we apply a local, compressed representation of the node neighborhood.

%These structures are complex, but we can make sense of them so long as we can load and analyze   %In order to make sense of such structures
%(Although we usually seek to remove as many such motifs as possible through graph normalization, to do so requires that we be able to load and traverse the graph.)
%Our objective is to allow dynamic modification of graphs even when they contain nodes with very high path depth.
%This requires an efficient

A key design objective, motivated by experience working with graphs built from high quality eukaryotic genome assemblies, is to support graphs with very high path depth.
Such motifs can occur in collapsed repeat structures generated by ambiguous sequence homology relationships in repeats found in the centromeres and other segmental duplications.
To meet these needs, we follow the approach of the dynamic Graph BWT (GBWT) \citep{Siren:2020}, and built a dynamic, compressed model of the paths.

>>>>

\begin{enumerate}
\item They are relatively sparse, with low average node degree.
\item They can be sorted, so that most edges go between nodes that are close together in the sort order (and identifier space).
\item Paths are similar to each other---in general, sequences from a species pangenome are similar enough to be easily alignable.
%We incorporate concepts first introduced in the dynamic version of the Graph BWT \citep{Siren:2020} (GBWT), a scalable implementation of the graph extension of the positional Burrows-Wheeler transform \citep{Durbin_2014}.
\end{enumerate}

These properties have already been used to build efficient dynamic variation graph data structures \citep{Eizenga_2020_BX}.
%Here, we build on this work and focus on issues that have arisen during our recent work on the Human Pangenome.
However, we take this work further, focusing on bottlenecks that arose during our work with high-quality assemblies that cover almost all parts of the genome \citep{Logsdon_2021,Nurk_2021}.
%A key design objective, motivated by experience working with graphs built from high quality eukaryotic genome assemblies, is to support graphs with very high path depth.
We find that it is necessary to support graphs that have regions of very high path depth.
Such motifs can occur in collapsed repeat structures generated by ambiguous sequence homology relationships in repeats found in the centromeres and other segmental duplications.
If we cannot process such regions, we have only two options---we can remove these regions, or we can leave them unaligned.
Neither of these solutions allows us to investigate their biological features.
To seamlessy represent such difficult regions, we follow the approach of the dynamic Graph BWT (GBWT) \citep{Siren:2020}, and built a node-centric, dynamic, compressed model of the paths.
